Should We Compute Feature Effects on Training or Validation Data?

### 2025-01-07 paper/paper.tex
paper: add feature effect error chain and error estimators

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

\begin{align}
    \text{MSE}(x_S; PDP_{f,S}, PDP_{\hat f,S})
    &= \mathbb{E}_F[{(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f,S}(x_S))}^2]\\
    &= \underbrace{{(PDP_f(x) - \mathbb{E}_F[PDP_{\hat{f}}(x)])}^2}_{Bias^2} + \underbrace{\text{Var}_F[PDP_{\hat{f}}(x)]}_{Variance}
\end{align}

\noindent Here $F$ denotes the distribution of trained models.

% estimation of PDP / ALE + additional MC error

$\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. In the simulation scenario, this yields:

\begin{align}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, PDP_{\hat f,S}) &= \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
\end{align}

% also for bias and variance (including m-1)

% estimation of MC error

% aggregation over x

@@

%
% ---- Bibliography ----
\bibliographystyle{splncs04}
\bibliography{paper/mybibliography}
\end{document}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{svg}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

\begin{equation}
    \text{MSE}(x_S; PDP_{f,S}, PDP_{\hat f,S})
    = \mathbb{E}_F[{(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f,S}(x_S))}^2]
\end{equation}
\begin{equation}
    = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f},S}(x_S)])}^2}_{Bias^2} + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance}
\end{equation}

\noindent Here, $F$ denotes the distribution of trained models. The bias
is linked to the bias of the model, the variance comes from the variance
in the model fits (randomness in training data, randomness in model training procedure).

Since we usually cannot determine $PDP_{\hat f,S}$, we need to estimate it by
Monte Carlo integration, yielding $\widehat{PDP}_{\hat f,S}(x_S)$. This,
however, introduces an additional variance term (we use the random variable
$X_{mc}$ for the Monte Carlo samples (e.g., training or validation data)):

\begin{equation}
    \begin{split}
        \text{MSE}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S})
        = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f}, S}(x_S)])}^2}_{Bias^2} \\
        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X_{mc}}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
    \end{split}
\end{equation}

\begin{proof}
    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:

    \begin{align*}
        \text{MSE}(PDP_{f}, \widehat{PDP}_{\hat f})
         & = \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_{\hat f})^2]                                                                                         \\
         & = \mathbb{E}_F \mathbb{E}_X[PDP_f^2 - 2PDP_f\widehat{PDP}_{\hat f} + \widehat{PDP}_{\hat f}^2]                                                         \\
         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_{\hat f}^2]                                                      \\
         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_{\hat f}]^2] \\
         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \text{Var}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])   \\
         & \quad + \mathbb{E}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])^2                                                                                           \\
         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] +  Var_F[PDP_{\hat f}] + \mathbb{E}_F[PDP_{\hat f}]^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]  \\
         & = (PDP_f - \mathbb{E}_F[PDP_{\hat f}])^2 + \text{Var}_F[PDP_{\hat f}] + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
    \end{align*}

    \noindent At multiple points, we use the fact that $\mathbb{E}_X[\widehat{PDP}_{\hat f}]
        = PDP_{\hat f}$ (cf.~\cite{molnar_relating_2023}).
\end{proof}

\noindent We see that the variance due to MC integration also depends on the model
distribution $F$. Similarly, one could use the estimate $\widehat{PDP}_{f,S}$
as groundtruth, introducing an additional variance term $Var_{X_{mc2}}$when
estimated on different MC sample (proof in
\textsc{Appendix~\ref{app:proof-mc-variance}}).
An overview of all error terms in this chian can be found in
\textbf{Fig.\@~\ref{fig:error-graph}}.\\

\begin{figure}[t]
    \includegraphics[width=\textwidth]{img/error_graph.pdf}
    \caption{Error chain in feature effect estimation
        (modified, original version can be found in~\cite{molnar_relating_2023})}\label{fig:error-graph}
\end{figure}

\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. We propose the following estimators:
\begin{equation}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
\end{equation}
\begin{equation}
    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
\end{equation}
\begin{equation}
    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
    \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
\end{equation}

\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
but we do not specify which data points to use for MC integration. The
variance captures both the variance in the model fits and the variance due to
MC integration. To estimate the MC variance, we propose the following estimators:
\begin{equation}
    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
\end{equation}

and
\begin{equation}
    \begin{split}
        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
        \left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)^2
    \end{split}
\end{equation}

\noindent For more convenient analysis of the errors, one could also
aggregate them over the marginal distribution of $X_S$ (e.g.,
estimated by averaging over the grid points if chosen appropriately)
to obtain a single error measure per feature effect.

\section{Overview of Test Functions}\label{sec:test-functions}

@@

%
% ---- Bibliography ----
\bibliographystyle{splncs04}
\bibliography{paper/mybibliography}

% appendix
\newpage
\appendix
\section{Proof: Monte Carlo Variance for Groundtruth}\label{app:proof-mc-variance}
\begin{proof}
    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:

    \begin{align*}
        \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_f)^2] & = \mathbb{E}_F\mathbb{E}_X[PDP_f^2 - 2PDP_f \widehat{PDP}_f + \widehat{PDP}_f^2]                                 \\
                                                              & = PDP_f^2 - 2PDP_f \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f^2]       \\
                                                              & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_f]^2] \\
                                                              & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + PDP_f^2                                       \\
                                                              & = \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f]                                                                      \\
                                                              & = \text{Var}_X[\widehat{PDP}_f]
    \end{align*}

    \noindent Again we use $\mathbb{E}_X[\widehat{PDP}_{f}] = PDP_{f}$ (cf.~\cite{molnar_relating_2023}) as well as the fact
    that all quantities based on $f$ do not depend on the model distribution $F$.
\end{proof}

\end{document}
